---
title: "Group Project"
author: "Your NAME goes here"
execute: 
  echo: false
  eval: true
number-sections: true
format:
  html:
    embed-resources: true
editor_options: 
  chunk_output_type: console
---

```{r}
#| warning: false
#| message: false
library(tidyverse)
library(ggplot2)
library(tidymodels)
library(sjPlot)
library(performance)
library(skimr)
library(kableExtra)
library(dplyr)
library(knitr)

obesity <- read.csv("DAProject1.csv")
obesity$Obese <- as.factor(obesity$Obese)
obesity$Year <- as.factor(obesity$Year)
```


```

# Introduction {#sec-intro}

# Exploratory Analysis {#sec-EA}

```{r}
#| label: fig-bar
#| fig-cap: "Obesity by Year Bar Plot"

ggplot(obesity, aes(x = Year, fill = Obese)) + 
  geom_bar(colour = "black", position = "fill") + 
  scale_fill_manual(values = c("Yes" = "lightblue", "No" = "salmon")) + 
  labs(y = "Proportion", 
       title = "Obesity Proportion by Year")
```

```{r}
#| label: tbl-prevelance
#| tbl-cap: "prevelance comparison table"

obesity %>%
  group_by(Year) %>%
  summarise(
    Obese = sum(Obese == "Yes"),
    Total = n(),
    Prevalence = Obese / Total
  ) %>%
  kable(digits = 3, caption = "Obesity Prevalence by Year")
```

# Formal Analysis {#sec-FA}

# Conclusions {#sec-con}

------------------------------------------------------------------------

# Further Questions {#sec-further}

## Part 1

```{r}
#| echo: true
#| tidy: true

# Data has been pre-loaded
q2_data <- read.csv(url('https://github.com/DA-STATS3011-STATS4052/DA-course_materials/raw/main/docs/data_sets/test1.csv'))
# Code for further Question 1 part 1

```

## Part 2

```{r}
#| echo: true
# Code for further Question 1 part 2

```




