---
title: An Analysis of Obesity Prevalence and Associated Factors in Scotland (2008–2012)
author: Group 1
execute: 
  echo: false
  eval: true
number-sections: true
format:
  html:
    embed-resources: true
editor_options: 
  chunk_output_type: console
---

```{r}
#| warning: false
#| message: false
library(tidyverse)
library(ggplot2)
library(tidymodels)
library(sjPlot)
library(performance)
library(skimr)
library(kableExtra)
library(dplyr)
library(knitr)
```

```{r}
obesity <- read.csv("DAProject1.csv")
obesity$Obese <- as.factor(obesity$Obese)
obesity$Year <- as.factor(obesity$Year)
```

# Introduction {#sec-intro}
Obesity is a major public health concern, associated with an increased risk of chronic conditions such as cardiovascular disease, type 2 diabetes, and certain cancers. Understanding how obesity prevalence changes over time, and how it is influenced by demographic, socio-economic, and lifestyle factors, is essential for informing effective health policy and interventions.
This report analyses data from the Scottish Health Surveys conducted between 2008 and 2012. The dataset includes individual-level information on age, sex, educational attainment, dietary behaviours (fruit and vegetable consumption), and obesity status. These variables provide a valuable opportunity to explore both temporal trends and potential risk factors associated with obesity within the Scottish population.
The primary aim of this analysis is to assess whether the prevalence of obesity in Scotland has changed over the period 2008–2012. In addition, the report investigates whether obesity varies across key demographic groups, socio-economic indicators, and lifestyle behaviours.

# Exploratory Analysis {#sec-EA}

```{r}
#| label: fig-bar
#| fig-cap: "Obesity by Year Bar Plot"

ggplot(obesity, aes(x = Year, fill = Obese)) + 
  geom_bar(colour = "black", position = "fill") + 
  scale_fill_manual(values = c("Yes" = "lightblue", "No" = "salmon")) + 
  labs(y = "Proportion", 
       title = "Obesity Proportion by Year")
```
The bars in the plot appear to be fairly consistent. This indicates that the prevalence of obesity across Scotland over the years in the survey has not changed much, although it may be more clear from a table.

```{r}
#| label: tbl-prevelance
#| tbl-cap: "prevelance comparison table"

obesity %>%
  group_by(Year) %>%
  summarise(
    Obese = sum(Obese == "Yes"),
    Total = n(),
    Prevalence = Obese / Total
  ) %>%
  kable(digits = 3, caption = "Obesity Prevalence by Year")
```
Now using a table we see that the prevalence of obesity is 0.296 in 2008, 2011 and 2012. It slightly dropped in 2009 and peaked in 2010 but not enough to infer that the prevalence changed over the years. Furthermore, these two outliers are still pretty close to 0.296.

# Formal Analysis {#sec-FA}

# Conclusions {#sec-con}




